Stateful Forward Reasoning

### 2025-02-05 paper.tex
Restore some newcommands that were accidentally deleted
This partially reverts commit 62d7792a5e283487598f09894f862964333ca330.
Fix algorithms
Introduction first draft

\section{Introduction}
\xcom{TODO}

\xcom{ Specify the problem (user defined rules)
 + deal with changes with local context}
\xcom{ Talk about aesop}
\xcom{ Do we outline the solution?}

An important contextualization is that the forward reasoning procedure in this paper takes place within Aesop's routine procedure.
We will focus on examples that highlight this style of proof while noting that in practice, a combination of both forwards and backwards reasoning are used.

@@

\begin{algorithm}
    \caption{New hypothesis}\label{alg:addhyp}
    \KwData{A hypothesis $h : T$ that matches with $T_i$.}
    \KwResult{The maps $\eta$ and $\mu$ are updated.}
        \eIf{$i = 1$}{
            \For{$x$ \Kwin $s(i)$}{
                Update $\mu_x$ with $(1,\sigma_m(x)) \mapsto \mu_x(1,\sigma_m(x))\cup\{[h]\}$ (A level 1 match).\\
            }
            Extend match $[h]$.
            }
        {
        \For{$x$ \Kwin $s(i)$}{
            Update $\eta_x$ with $(i,\sigma_h(x)) \mapsto \eta_x(i,\sigma_h(x))\cup\{h\}$.\\        
        }    
        $ms$ \gets \ $\bigcap_{x \in s(i)} \mu_x \left(i - 1, \sigma_h(x)\right)$\;
        \For{$m$ \Kwin $ms$}{
            $m'$ \gets m.\text{append } h\\
            \For{$x$ \Kwin $s(i)$}{
            Update $\mu_x$ with $(i,\sigma_m(x)) \mapsto \mu_x(i,\sigma_m(x))\cup\{m'\}$.}\\
            Extend match $m'$.
        }}
\end{algorithm}

\begin{algorithm}
    \caption{Extend match}\label{alg:updatematch}
    \KwData{A partial match $m$ of level $i$.}
    \KwResult{The map $\mu$ is updated.}  
        $hs$ \gets \ $\bigcap_{x \in s(i + 1)} \eta_x \left(i + 1, \sigma_m(x)\right)$\\
        \For{$h$ \Kwin $hs$}{
            $m'$ \gets m.\text{append } h\\
            \For{$x$ \Kwin $s(i + 1)$}{
            Update $\mu_x$ with $(i + 1,\sigma_{m'}(x)) \mapsto \mu_x(i + 1,\sigma_{m'}(x))\cup\{m'\}$.}\\
            Extend match $m'$.
        }
\end{algorithm}

>>>>

\section{Introduction}

In many proof assistants, simple automation tactics based on tree search play a substantial role in practice: Isabelle's (TODO ref) and Rocq's (TODO ref) auto, ACL2's (TODO ref) waterfall, PVS's (TODO ref) grind, Lean's (TODO ref) Aesop (TODO ref), etc.
These tactics operate on \emph{goals} $Γ ⊢ A$, where the \emph{local context} $Γ$ is a list of assumptions and $A$ is the proposition to be proved.
Faced with an initial goal, the tactics try various user-specified \emph{rules}, which correspond to admissible inferences in the prover's logic.
A successful rule reduces the current goal to zero or more subgoals, which are solved recursively until either a full proof is found or no rule applies to any open goal or the search times out.

Typical rules would be the $∧$-introduction lemma $A → B → A ∧ B$, which, when applied backwards, reduces a goal $Γ ⊢ A ∧ B$ to subgoals $Γ ⊢ A$ and $Γ ⊢ B$; or left $∨$-introduction, which reduces $Γ ⊢ A ∨ B$ to $Γ ⊢ A$.
Such \emph{backward rules} are the most common rule type, but most search tactics allow users to use any tactic as a rule.
These rule tactics can output arbitrary subgoals, subject only to the condition that proofs of the returned subgoals can be assembled into a proof of the goal on which the rule was run.

ACL2 and Aesop additionally provide special support for \emph{forward rules}.
Given, for example, the transitivity lemma $\All{x~y~z}{x < y → y < z → x < z}$, a forward rule run on the goal $a < b,\, b < c ⊢ A$ produces the single subgoal $a < b,\, b < c,\, a < c ⊢ A$.
Hence forward rules can be used to establish additional facts that may be relevant for other rules.
For example, the Aeneas project (TODO ref) uses Aesop's forward rules \jcom{(ensure that this is accurate)} in its proof automation to establish facts that are later used by an arithmetic decision procedure.

However, Aesop's (and ACL2's) support for forward rules was, until recently, quite basic.
In particular, each forward rule was tried on each goal separately (subject to pre-filtering by a discrimination tree (TODO ref) index).
This is wasteful since in practice, many rules produce subgoals with a similar---or even the same---local context as the initial goal.
For example, the ∧-introduction rule affects only the goal's target, not its context.
Still, all forward rules applied to the initial $Γ ⊢ A ∧ B$ would be tried again on the subgoals $Γ ⊢ A$ and $Γ ⊢ B$, with predictably little effect.

In this paper, we present an alternative way to apply forward rules that addresses this issue.
The central idea is to remember, for each goal $Γ ⊢ A$, which forward rules can be partially applied to which sets of hypotheses.
When a non-forward rule then generates subgoals $Δ₁ ⊢ A₁$, \dots, $Δₙ ⊢ Aₙ$, the new local contexts $Δᵢ$ are likely similar to $Γ$.
Hence it is generally cheaper (and often much cheaper) to start with the partial matches for $Γ$ and update them for each of the $Δᵢ$.
The update first deletes partial matches that include hypotheses no longer present in $Δᵢ$.
Afterwards it adds new partial matches for the hypotheses that are new in $Δᵢ$, extending the existing partial matches for a rule if the new hypothesis is compatible with them.
Whenever this process produces a full match---i.e., all premises of a forward rule are satisfied by a set of hypothesis in $Δᵢ$---the rule is applied and adds a corresponding new hypothesis to $Δᵢ$.
Our main contribution is a data structure, the \emph{partial match state}, that organises the partial matches in such a way that the update can be done efficiently.

In addition to supporting such \emph{incremental} forward reasoning, the partial match state satisfies two other desiderata:
\begin{itemize}
  \item It is \emph{persistent}, i.e.\ partial match states for parent and child goals share much of their structure.
        This is desirable since Aesop supports search strategies other than depth-first search and must therefore keep all goals encountered during the search, along with their partial match states, in memory.
  \item It reduces the number of \emph{unifications} performed when matching hypotheses against rule premises.
        In dependent type theory, unification happens up to definitional equality, which makes it quite expensive.
        We therefore want to avoid redundant unifications.
\end{itemize}

Our evaluation (Sec.~\ref{sec:evaluation}) shows that this new implementation of forward reasoning in Aesop substantially outperforms the previous, naive implementation in some (though not all) cases.
The implementation, the benchmarks we used and the full evaluation data are available in a supplement to this paper (TODO ref).
Mainline Aesop already uses the new implementation by default.

@@

\begin{algorithm}
    \caption{New hypothesis}\label{alg:addhyp}
    \KwData{A hypothesis $h : T$ that matches with $T_i$.}
    \KwResult{The maps $\eta$ and $\mu$ are updated.}
    \eIf{$i = 1$}{
      \For{$x$ \Kwin $s(i)$}{
        Update $\mu_x$ with $(1,\sigma_m(x)) \mapsto \mu_x(1,\sigma_m(x))\cup\{[h]\}$ (a level 1 match)
      }
      Extend match $[h]$
    }{
      \For{$x$ \Kwin $s(i)$}{
        Update $\eta_x$ with $(i,\sigma_h(x)) \mapsto \eta_x(i,\sigma_h(x))\cup\{h\}$
      }
      $ms \gets \bigcap_{x \in s(i)} \mu_x \left(i - 1, \sigma_h(x)\right)$ \\
      \For{$m$ \Kwin $ms$}{
        $m' \gets m.\text{append } h$ \\
        \For{$x$ \Kwin $s(i)$}{
          Update $\mu_x$ with $(i,\sigma_m(x)) \mapsto \mu_x(i,\sigma_m(x))\cup\{m'\}$
        }
        Extend match $m'$
      }
    }
\end{algorithm}

\begin{algorithm}
    \caption{Extend match}\label{alg:updatematch}
    \KwData{A partial match $m$ of level $i$.}
    \KwResult{The map $\mu$ is updated.}  
        $hs \gets \bigcap_{x \in s(i + 1)} \eta_x \left(i + 1, \sigma_m(x)\right)$ \\
        \For{$h$ \Kwin $hs$}{
            $m' \gets m.\text{append } h$\\
            \For{$x$ \Kwin $s(i + 1)$}{%
              Update $\mu_x$ with $(i + 1,\sigma_{m'}(x)) \mapsto \mu_x(i + 1,\sigma_{m'}(x))\cup\{m'\}$.%
            }
            Extend match $m'$.
        }
\end{algorithm}
